Distinct smell and taste disorder phenotype of post-acute COVID-19 sequelae

### 2022-06-02 paper/markdown/figures_and_tables.Rmd
Major analysis update
Corresponds now to the analyses presented in the pre-print

---
title: "Hyposmia as a standalone persistent symptom of long COVID"
subtitle: "Figures and Tables"
author: "Health after COVID-19 in Tyrol study team"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  pdf_document:
    latex_engine: xelatex
bibliography: cov_biblio.bib
csl: american-medical-association.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dev = "cairo_pdf")

@@

# Tables {#sec:tbl}

```{r tab-covild}

paper_tables$cohort_covild %>% 
  format_line %>%  
  kable(caption = '\\textbf{Characteristic of the CovILD study cohort.}\\\\Ambulatory, Moderate, Severe: severity of acute SARS-CoV-2 infection, Test: statistical test used for the comparison between the severity strata (KW: Kruskal-Wallis), Significance: test p value corrected for multiple comparisons with Benjamini-Hochberg method.', 
        format = 'latex', 
        escape = F, 
        booktabs = T, 
        longtable = T) %>% 
  column_spec(1, width = '3cm') %>% 
  column_spec(2:4, width = '3cm') %>% 
  column_spec(5, width = '0.6cm') %>% 
  column_spec(6, width = '2cm') %>% 
  collapse_rows(columns = 1) %>% 
  kable_styling(full_width = F, 
                latex_options = c('striped', 
                                  'hold_position', 
                                  'repeat_header'))

@@

```{r tab-survey}

paper_tables$cohort_survey %>% 
  format_line %>%  
  kable(caption = '\\textbf{Characteristic of the Health after COVID-19 in Tyrol study cohorts.}\\\\Test: statistical test used for the comparison between the severity strata (U: Mann-Whitney U test), Significance: test p value corrected for multiple comparisons with Benjamini-Hochberg method.', 
        format = 'latex', 
        escape = F, 
        booktabs = T, 
        longtable = T) %>% 
  column_spec(1, width = '3.5cm') %>% 
  column_spec(2:3, width = '4cm') %>% 
  column_spec(4, width = '0.6cm') %>% 
  column_spec(5, width = '2cm') %>% 
  collapse_rows(columns = 1) %>% 
  kable_styling(full_width = F, 
                latex_options = c('striped', 
                                  'hold_position', 
                                  'repeat_header'))

@@

```{r fig-consort, fig.width = mm_inch(paper_figures$consort$w), fig.height = mm_inch(paper_figures$consort$h), dpi = 600, fig.cap = 'CONSORT flow diagrams for the CoVILD and HACT study cohorts.'}

paper_figures$consort$plot

@@

__Figure \@ref(fig:fig-consort). CONSORT flow diagrams for the CoVILD and HACT study cohorts.__

@@

```{r fig-kinetic, fig.width = mm_inch(paper_figures$kinetic_modeling$w), fig.height = mm_inch(paper_figures$kinetic_modeling$h), dpi = 600, fig.cap = 'Second order kinetic modeling of symptom recovery course in the HACT study.'}

paper_figures$kinetic_modeling$plot

@@

__Figure \@ref(fig:fig-kinetic). Second order kinetic modeling of symptom recovery course in the HACT study.__

Symptom frequency during acute COVID-19 and recovery in the Austria and Italy HACT cohorts (0, 7, 14, 28 and 90 days post symptom onset) were modeled with mixed-effect logistic regression (__Supplementary Table S2__). Significance of the first and second order model terms was determined by step-wise likelihood ratio test (LRT). P values were corrected for multiple comparisons with Benjamini-Hochberg method. N number of complete observations is provided under the plots in (__B__). Dim.: diminished, imp.: impaired.

__(A)__ Significance of the first and second order model term in LRT test. Points represent single symptoms, color codes for significance (gray: non-significant). Top 10 symptoms with the most significant second order terms as candidate long-term persistent features are labeled with their names.

__(B)__ Median recovery times for top 10 symptoms with the most significant second order terms as candidate long-term persistent features. Whiskers represent interquartile ranges.

@@

```{r fig-hyposmia-kinetic, fig.width = mm_inch(paper_figures$hyposmia_kinetic$w), fig.height = mm_inch(paper_figures$hyposmia_kinetic$h), dpi = 600, fig.cap = 'Actual and predicted frequency of self-reported hyposmia in course of COVID-19 recovery.'}

paper_figures$hyposmia_kinetic$plot

@@

__Figure \@ref(fig:fig-hyposmia-kinetic). Actual and predicted frequency of self-reported hyposmia in course of COVID-19 recovery.__

Frequency of self-reported hyposmia during acute COVID-19 and recovery in the HACT study (__A__, 0, 7, 14, 28 and 90 days post symptom onset) and severity subsets of the CovILD cohorts (__B__, acute COVID: 0, 60-, 100 and 180-day follow-up visits) was modeled with mixed-effect logistic regression (__Supplementary Table S2__). Points and solid lines represent the actual frequencies. The model predictions are displayed as dashed lines. The ranges of complete observations per time point are shown under the plots.

@@

```{r fig-hyposmia-overlap, fig.width = mm_inch(paper_figures$hyposmia_overlap$w), fig.height = mm_inch(paper_figures$hyposmia_overlap$h), dpi = 600, fig.cap = 'Co-occurrence of self-reported hyposmia and other symptoms in course of COVID-19 recovery.'}

paper_figures$hyposmia_overlap$plot

@@

__Figure \@ref(fig:fig-hyposmia-overlap). Co-occurrence of self-reported hyposmia and other symptoms in course of COVID-19 recovery.__

Cosine similarity coefficients between hyposmia and other self-reported symptoms of COVID-19 in the HACT 
(__A__, `r length(hypo_overlap$variables$survey) + 1` symptoms in total, Austria: n = `r hypo_overlap$analysis_tbl$survey$survey_at$ID %>% unique %>% length`, Italy: `r hypo_overlap$analysis_tbl$survey$survey_it$ID %>% unique %>% length`) 
and COVILD study 
(__B__, `r length(hypo_overlap$variables$covild) + 1` symptoms, ambulatory: n = `r hypo_overlap$analysis_tbl$covild$A$ID %>% unique %>% length`, moderate: n = `r hypo_overlap$analysis_tbl$covild$HM$ID %>% unique %>% length`, severe COVID-19: `r hypo_overlap$analysis_tbl$covild$HS$ID %>% unique %>% length`) were calculated for the indicated time points (__Supplementary Table S3__) and presented in radial plots. Each bar represents a single symptom. Bar length and color code for the value of Cosine similarity coefficient. Dashed lines represent cosine similarity coefficient of 0.5.

@@

```{r fig-clustering, fig.width = mm_inch(paper_figures$clustering$w), fig.height = mm_inch(paper_figures$clustering$h), dpi = 600, fig.cap = 'Long-term COVID-19 recovery phenotypes in the HACT study.'}

paper_figures$clustering$plot

@@

__Figure \@ref(fig:fig-clustering). Long-term COVID-19 recovery phenotypes in the HACT study.__

HACT participants with COVID-19 symptoms present for at least three months post clinical onset (Austria: n = `r nrow(pheno$analysis_tbl$survey_at)`, Italy: n = `r nrow(pheno$analysis_tbl$survey_it)`) were clustered in respect to manifestations present in at least 25% of the cohort for 28 days or longer with a combined algorithm employing self-organizing map (`r pheno$som_grid$xdim` $\times$ `r pheno$som_grid$ydim` hexagonal units, Jaccard distance) and hierarchical clustering (Ward D2 algorithm, Euclidean distance). Numbers of individuals assigned to particular clusters are presented under the plot in (__A__). Imp.: impaired.

(__A__) Presence of the features used for cluster definition in the participants assigned to the 'non-specific', 'hyposmia' and 'neuro/fatigue' subset.

(__B__) Frequency of self-reported hyposmia, taste disorders and tiredness at day in the participant subsets. Statistical significance of the frequency differences was assessed with $\chi^2$ test corrected for multiple comparisons with Benjamini-Hochberg method. P values are presented in the plot captions.

@@

```{r fig-risk, fig.width = mm_inch(paper_figures$hyposmia_risk$w), fig.height = mm_inch(paper_figures$hyposmia_risk$h), dpi = 600, fig.cap = 'Risk factors for long-term persistent self-reported post-COVID-19 hyposmia.'}

paper_figures$hyposmia_risk$plot

@@

__Figure \@ref(fig:fig-risk). Risk factors for long-term persistent self-reported post-COVID-19 hyposmia.__

Association of candidate risk factors (__Supplementary Table S1__) with long-term persistent self-reported hyposmia in the CovILD (__A__, hyposmia declared at the 100-day follow-up visit) and HACT cohorts (__B__, self-reported hyposmia present for at least three months post clinical onset) was investigated with univariable logistic regression (__Suppmenetary Table S4__). For the HACT cohorts, the models were weighted for age and sex distribution in the general COVID-19 convalescent populations and included stratified the observation time variable as a confounder. Significance of model estimates (odds ratio, OR) was determined by Wald Z test corrected for multiple comparisons with Benjamini-Hochberg method. OR estimate values with 95% confidence intervals for significant and nearly significant (adjusted p < 0.1) co-variates are presented as points and whiskers in the Forest plots. The ranges of complete observations per time point are shown under the plots.

>>>>

---
title: "Distinct smell and taste disorder phenotype of post-acute COVID-19 sequelae"
subtitle: "Tables and figures"
author: "HACT and CovILD study teams"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: hyposmia_biblio.bib
csl: frontiers_medical.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dev = "png", dpi = 600)

@@

# Tables

```{r tab-hact, tab.cap = 'Baseline characteristic of the HACT study Austria (AT) and Italy (IT) cohorts.'}

paper_tbl$hact_baseline %>% 
  flextable %>% 
  width(1, width = 3.7, unit = 'cm') %>% 
  width(2:3, width = 4, unit = 'cm') %>% 
  width(4, width = 2.6, unit = 'cm') %>% 
  width(5, width = 2.2, unit = 'cm') %>% 
  footnote(1, 4:5, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Mann-Whitney U test with wilcoxon r effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('BMI: body mass index, overweight > 25 kg/m\u00B2, obesity > 30 kg/m\u00B2,'), ref_symbols = 'b', part = 'body') %>% 
  footnote(15, 1, value = as_paragraph('Frequent respiratory infections, > 2 per year.'), ref_symbols = 'c', part = 'body') %>% 
  footnote(16, 1, value = as_paragraph('Frequent bacterial infections with antibiotic therapy, > 2 per year.'), ref_symbols = 'd', part = 'body') %>% 
  theme_vanilla %>%
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

@@

```{r tab-covild, tab.cap = 'Baseline characteristic of the CovILD study cohort and the study participants stratified by COVID-19 severity.'}

paper_tbl$covild_baseline %>% 
  flextable %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2:5, width = 2.7, unit = 'cm') %>% 
  width(6, width = 2.6, unit = 'cm') %>% 
  width(7, width = 2.2, unit = 'cm') %>% 
  footnote(1, 6:7, value = as_paragraph('Comparison of ambulatory, moderate and severe COVID-19 individuals. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('BMI: body mass index, overweight > 25 kg/m\u00B2, obesity > 30 kg/m\u00B2,'), ref_symbols = 'b', part = 'body') %>% 
  theme_vanilla %>%
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

@@

```{r fig-consort, fig.width = mm_inch(figures$consort$w), fig.height = mm_inch(figures$consort$h), fig.cap = 'Flow diagram of the analysis inclusion process for the CovILD and HACT studies.'}

figures$consort$plot

@@

__Figure \@ref(fig:fig-consort). Flow diagram of the analysis inclusion process for the CovILD and HACT studies..__

@@

```{r fig-sympt-recovery, fig.width = mm_inch(figures$sympt_recovery$w), fig.height = mm_inch(figures$sympt_recovery$h), fig.cap = 'Symptom-specific recovery times in ambulatory COVID-19.'}

figures$sympt_recovery$plot

@@

__Figure \@ref(fig:fig-sympt-recovery). Symptom-specific recovery times in ambulatory COVID-19.__

Symptom-specific recovery times were calculated for each participants of the HACT study cohorts (Austria: AT, Italy: IT).

__(A)__ Distribution of the recovery times in the individuals with the indicated symptoms present during acute COVID-19. Diamonds represent median recovery times, tinted ellipses code for interquartile ranges. Numbers of complete observations are indicated in the plot captions.

__(B)__ Percentages of individuals with smell and taste disorders in the AT (Austria) and IT (Italy) HACT study cohorts at particular time points after clinical onset. Numbers of complete observations are indicated under the plots.

@@

```{r fig-sympt-mds, fig.width = mm_inch(figures$sympt_mds$w), fig.height = mm_inch(figures$sympt_mds$h), fig.cap = 'Self-reported smell and taste disorders are isolated symptoms of long COVID and PASC.'}

figures$sympt_mds$plot

@@

__Figure \@ref(fig:fig-sympt-mds). Self-reported smell and taste disorders are isolated symptoms of long COVID and PASC.__

Binary symptom occurrence data during for long COVID ($\geq$ 28 days after clinical onset, __A__) and PASC ($\geq$ 90 days, __B__) in the HACT Austria (AT) and Italy (IT) cohorts were subjected to two-dimensional multi-dimensional scaling (MDS) with simple matching distance (SMD) between the symptoms. MDS coordinates are presented in point plots. Selected data points are labeled with the symptom names. Percentages of the data set variance associated with the MDS dimensions are indicated in the plot axes. Numbers of complete observations are indicated under the plots.

@@

```{r fig-apriori, fig.width = mm_inch(figures$apriori$w), fig.height = mm_inch(figures$apriori$h), fig.cap = 'Frequent co-occurrence of smell and taste disorders in long COVID and PASC.'}

figures$apriori$plot

@@

__Figure \@ref(fig:fig-apriori). Frequent co-occurrence of smell and taste disorders in long COVID and PASC.__

Frequent combinations of symptoms of long COVID (__A__) and PASC (__B__) in the HACT study Austria (AT) and Italy (IT) cohorts were identified with the apriori algorithm. Symptom combination support and confidence are presented in the plots. Point color codes for lift values. The hypo/anosmia and hypo/ageusia combinations are labeled.

@@

```{r fig-clustering, fig.width = mm_inch(figures$clustering$w), fig.height = mm_inch(figures$clustering$h), fig.cap = 'Clustering of ambulatory COVID-19 individuals by symptom-specific recovery times.'}

figures$clustering$plot

@@

__Figure \@ref(fig:fig-clustering). Clustering of ambulatory COVID-19 individuals by symptom-specific recovery times.__

Individuals of the training Austria (AT) cohort of the HACT study were subjected to clustering in respect to symptom-specific recovery times with the PAM (partitioning around medoids) algorithm and Euclidean distance measure (__Supplementary Figure S8__). Cluster assignment in the test Italy (IT) HACT cohort was done with k-NN label propagation algorithm. Recovery times for particular COVID-19 symptoms in the COVID-19 recovery clusters are presented as heat maps. Numbers of individuals assigned to the recovery clusters are indicated under the plots.

Dim. appetite: diminished appetite, Imp. concentration: impaired concentration, Imp. walk: impaired walk, Imp. FMS: impaired fine motor skills.

@@

```{r fig-sympt-clusters, fig.width = mm_inch(figures$sympt_clusters$w), fig.height = mm_inch(figures$sympt_clusters$h), fig.cap = 'Duration of neurocognitive and respiratory symptoms, fatigue, smell and taste disorders differs between the COVID-19 recovery clusters.'}

figures$sympt_clusters$plot

@@

__Figure \@ref(fig:fig-sympt-clusters). Duration of neurocognitive and respiratory symptoms, fatigue, smell and taste disorders differs between the COVID-19 recovery clusters.__

Semi-supervised clustering was performed as presented in __Figure \@ref(fig:fig-clustering)__ and __Supplementary Figure S8__. Mean recovery times in the recovery clusters are presented as lines, 2 $\times$ SEM intervals are displayed as tinted regions.  Numbers of individuals assigned to the recovery clusters are indicated under the plots.

Dim. appetite: diminished appetite, Imp. concentration: impaired concentration, Imp. walk: impaired walk, Imp. FMS: impaired fine motor skills.

```{r fig-base-clusters, fig.width = mm_inch(figures$base_clusters$w), fig.height = mm_inch(figures$base_clusters$h), fig.cap = 'COVID-19 recovery clusters differ in sex distribution, comorbidity and daily medication rates.'}

figures$base_clusters$plot

__Figure \@ref(fig:fig-base-clusters). COVID-19 recovery clusters differ in sex distribution, comorbidity and daily medication rates.__

Differences in sex distribution (__A__), frequency of comorbidity (__B__) and daily medication (__C__) between the recovery clusters in the training Austria (AT) and the test Italy (IT) HACT study cohorts (__Figure \@ref(fig:fig-clustering)__, __Supplementary Figure S8__) were assessed by $\chi^2$ test with Cramer V effect size statistic. P values were corrected for multiple testing with Benjamini-Hochberg method. The frequencies are presented as bar plots. Effect size statistics and p values are indicated in the plot caption. Numbers of complete observations are displayed under the plots.

```{r fig-recovery-clusters, fig.width = mm_inch(figures$recovery_clusters$w), fig.height = mm_inch(figures$recovery_clusters$h), fig.cap = 'Good clinical and mental health status and quality of life at the survey completion in individuals with persistent smell and taste disorders.'}

figures$recovery_clusters$plot

__Figure \@ref(fig:fig-recovery-clusters). Good clinical and mental health status and quality of life at the survey completion in individuals with persistent smell and taste disorders.__

Differences in minimum/maximum scaled readouts of clinical and physical recovery, mental health and quality of life scoring at the time of survey completion in the recovery clusters (__Figure \@ref(fig:fig-clustering)__, __Supplementary Figure S8__) are presented in radial plots. Yes/no items (incomplete convalescence, weight loss, new medication and need for rehabilitation) were binarized (yes: 1, no: 0) prior to visualization. Lines represent mean values, 2 $\times$ SEM intervals are displayed as tinted regions. Numbers of individuals assigned to the recovery clusters are indicated under the plots.

incomplete recovery: self-reported incomplete recovery, # persist. symptoms: number of long COVID symptoms ($\geq$ 28 days after clinical onset), phys. performance loss: physical performance loss as compared with the time before COVID-19, QoL score: quality of life impairment score, OMH score: overall mental health impairment score, ANX score: anxiety score, DPR: depression score.
